---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A1. There are 32 departments that have offered climate courses with at least one seat. Atmospheric and Oceanic Sciences have the most courses with 93.**

```{r}
climate_courses |> 
    filter(str_detect(title, "Climate") & seats >= 1) |>
  group_by(department)
  
```

```{r}
climate_courses |> 
  group_by(department) |>
  summarise(count = n()) |>
  
  arrange(desc(count))
```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? **A2. It's interesting that Engineering has a good number of climate courses. Hopefully, they are courses on how to build in ways that are energy efficient.**

**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3. Business Management offered the most courses with 174. As for syllabi, 134 departments have at least one posted. The biggest difference in the two top fives is that the engineering professional master's has a lot of syllabi posted with very few courses. Conversely, English offers a lot of courses but has a relatively low number of syllabi posted.**

```{r}
fall_courses <- umd_courses |> 
  filter(term == "202408")
fall_courses |>
  group_by(department) |>
  summarise(
    count_courses = n()
  ) |>
  
  arrange(desc(count_courses))
```

```{r}
fall_courses |>
  filter(syllabus_count >= 1) |>
  group_by(department) |>
  summarise(
    count_courses = n()
  ) |>
  
  arrange(desc(count_courses))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time. **A4. The data shows that spending by Wes Moore's committee increased drastically around the primary and general elections.**

```{r}
new_maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

head(maryland_expenses)
```

```{r}
maryland_expenses |>
   filter(str_detect(committee_name, "Moore"))
```

```{r}
maryland_expenses |>
  filter(str_detect(committee_name, "Moore")) |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(month))
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe). **A5. The data shows that a considerable amount more money was spent in Maryland than all other states combined, which is fascinating but unsurprising.**

```{r}
maryland_expenses_with_location <- maryland_expenses |>
  mutate(
    spending_location = case_when(
        str_detect(address, " Maryland ") ~ "In-state",
        .default = "Out of state"
    )
  )
```

```{r}
maryland_expenses_with_location |>
  filter(str_detect(amount, ".")) |>
  group_by(spending_location) |>
  summarize(total = sum(amount))
```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A6. In my pre lab, I asked how much was spent on office supplies. The answer is \$233,865.**

```{r}
maryland_expenses_supplies <- maryland_expenses |>
  filter(str_detect(expense_purpose, "Office Supplies"))
```

```{r}
maryland_expenses_supplies |>
  summarize(total = sum(amount))
```
