---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

# Load janitor
library (tidyverse)
library (janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

# Load in the Maryland grants and loan data
```

```{r}
cleaned_payments <- payments |>
  clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: I was able to find 70 records with five different spellings.

```{r}
library(tidyverse)
library(lubridate)
```

```{r}
pg_schools <- cleaned_payments |>
  filter (str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOL") | str_detect(vendor_name, "PRINCE GEORGE'S COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL")) |>
  group_by(vendor_name)
```

```{r}
# Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

#A2: 2009 had the most with eight. As time went on, payments trended downward with the lowest amount in the years leading into pandemic, which I thought was noteworthy.

pg_schools |>
  group_by(fiscal_year) |>
    summarise(fiscal_payments = n())
```

```{r}
cleaned_umd_payments |>
  clean_names() |>
  filter(str_detect(agency_name, "UNIVERSITY OF MARYLAND") & str_detect(agency_name, "CAPITAL")) |>
  group_by(vendor_name) |>
  summarise(vendor_amount = sum(amount)) |>
  arrange(desc(vendor_amount))
# Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

# Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

#A3. Some vendors (like Whiting Turner, Plano-Coudon and Precision X-Ray) clearly have duplicate entries that should be combined. To fix this, I would use get_dupes to look for duplicate payment amounts. I would then go in manually and either delete the duplicates (if they are double entries) or make them uniform with the other entries (if the identical payments are intentional).
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: On the whole, it seems as though stem cell research funding has trended down, but there was a recent surge in 2023. Most of the top years are from 15 years ago or so, aside from 2023. The top year, 2009, is a huge outlier for some reason. It could have something to do with the Obama administration taking over.

```{r}
md_stem_cells |>
  filter(str_detect(source, "STEM CELL") | str_detect(description, "STEM CELL")) |>
  group_by(fiscal_year) |>
  summarise(year_amount = sum(amount)) |>
  arrange(desc(year_amount))
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are multiple entries where the money is going to UMD. Some entries are more specific, like which department at UMD, and some only generally state the money is going to the university. There are also duplicates like "UNIVERSITY OF MARYLAND" and "UNIVERSITY OF MARYLAND COLLEGE PARK". To clean this data, I would first get rid of duplicates by fixing issues like the one I just mentioned. Then, I would make all mentions of UMD uniform but just having them say "UMD", so no more "UNIV OF MD", "UNIVERSITY OF MARYLAND", etc. Finally, I think I would have two different tables using this data: one of them would keep the extra details of saying where in the university the money is going when applicable, and the other would more simply lump all of the UMD entries together to see the total amount of money that went to the university. I think the most interesting entry here is the money going to the performing arts center. I'd to know what exactly that money funded.

```{r}
umd_zip_grants |>
  filter(str_detect(zip_code, "20742")) |>
  group_by(grantee) |>
  summarise(grantee_amount = sum(amount)) |>
  arrange(desc(grantee_amount))
```
