---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

**A1:** The data is being pulled from a website, not a local file.

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:** There are 9,134 rows, each with data on an earthquake in the U.S. The data includes a lot of location info on the earthquake (coordinates, place, locationSource). It shows the magnitude of the earthquake, when it happened and some other data that I'm not sure how to decipher.

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes |>
  arrange(desc(depth))
```

**A3: I arranged the entries by depth in descending order. It is immediately clear that the Fiji region has the deepest earthquakes. Argentina, Indonesia and the Philippines are the only other places near the top. The deepest earthquake was 643 kilometers deep in the Fiji region with a magnitude of 4.2. Virtually all of the Fiji earthquakes are around this magnitude.**

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
earthquakes |>
  filter(mag > 6.0)
```

**A4: I used a filter function to remove the smaller earthquakes. There are 12 earthquakes that meet the criteria.**

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
earthquakes |>
  filter(mag > 6.0 & depth < 20.0)
```

**A5: I added an extra condition to the filter using an ampersand, so the entries must meet both criteria. There are eight entries that meet both requirements.**

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
earthquakes |>
  filter(mag > 6.0 | depth < 20.0)
```

**A6: There are now 7,605 entries that meet at least one of the criteria, because there are many earthquakes with a depth less than 20km.**

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
earthquakes |>
  filter(str_detect(place, 'Alaska') | str_detect(place, 'AK'))
```

**A7: I filtered for "Alaska" and "AK" in the place column to find the entries I was looking for. There appear to be 2,353 earthquakes in Alaska in the dataset.**

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(count = n())
```

**A8: I found each type by grouping by type and counting the entries to get a simplified view. Earthquakes and quarry blasts are the two most common, but earthquakes are by far the top entry. Explosions are not far behind quarry blasts, and there were one each of rock slides and landslides.**

------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(count = n(),
            avg_depth = mean(depth))
```

**A9: I found the average depth by using the same code from the last question and adding a mean function for the depth of each type. The average earthquake depth is about 18.5km. It is interesting but logical that the explosions and quarry blasts would be above the wherever the earthquakes were monitored. However, I think its more interesting that the one rock slide was below the monitoring site. I'm curious if this monitoring site, wherever it was high above sea level.**

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes |>
  mutate(hour = hour(time)) |>
  group_by(hour) |>
  summarise(hour_count = n()) |>

arrange(desc(hour_count))
```

**A10: I extracted the hour by using a mutate function to make a new column with the hour of each event. Surprisingly, there actually is kind of an answer to this question. Most earthquakes seem to happen from 8am through 10am. I would love to know if this is just a coincidence or if there is science behind this.**

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}
earthquakes |>
  mutate(serious = case_when(
    mag > 3.0 ~ 'Y',
    mag < 3.0 ~ 'N',
    mag == 3.0 ~ 'N' 
  ))
```

**A11: I used a mutate/case_when function to create a "serious" column where there will be a Y if it is over 3.0 and an N if it is below. I also added an extra line for the entries that are exactly 3.0, marking them as N because they are not greater than 3.0.**

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquakes |>
  filter(mag > 3.0) |>
  mutate(hour = hour(time)) |>
    group_by(hour) |>
    summarise(hour_count = n()) |>
    
arrange(desc(hour_count))
```

**A12**: The above code is filtering out the earthquakes with lower magnitudes, extracting the hour each earthquake occurred and organizing them by hour based on how many serious earthquakes occurred then. The 5 a.m. hour had the fewest and the 8 p.m. hour had the most.

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?

```{r}
rockslide <- earthquakes |>
  filter(type == 'rockslide')
```

\^\^\^ What are the details of the one rock slide? Where was it?

The rock slide was 67km northeast of Nikolski, Alaska, a small town on a far-off Alaskan island. The islands in the area do have some elevation, so its possible the monitoring site was above the rock slide.

```{r}
earthquakes |>
  filter(type == 'explosion') |>
  group_by(locationSource) |>
  summarise(location_count = n()) |>
  arrange(desc(location_count))
```

\^\^\^ Where are explosions being registered? Are they demolitions, war or something else?

Seems like there are a lot of explosions in Washington state, which I did not expect. Maybe they are geothermal?

```{r}
earthquakes |>
  filter(type == 'explosion') |>
  summarise(avg_explosion = mean(mag))
```

```{r}
earthquakes |>
  filter(type == 'explosion') |>
  arrange(desc(mag))
```

\^\^\^ What is average magnitude of these explosions? What was the largest explosion

The average explosion has a magnitude of about 1.6. The largest explosion was 2.37 in Washington.
